---
title: "Landscape Metric Covariates"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Landscape Metric Covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(multiScaleR)
library(terra)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4.25,
  warning = FALSE,
  message = FALSE
)

pkg_extdata <- function(...) {
  src_path <- normalizePath(file.path("..", "inst", "extdata", ...),
                            winslash = "/",
                            mustWork = FALSE)
  if (file.exists(src_path)) {
    return(src_path)
  }

  inst_path <- system.file("extdata", ..., package = "multiScaleR")
  if (nzchar(inst_path) && file.exists(inst_path)) {
    return(inst_path)
  }

  stop("Could not locate required extdata file: ", paste(..., collapse = "/"))
}
```

# Why explicit covariate specifications?

Traditional `multiScaleR` workflows assume that each raster layer becomes one
optimized, kernel-weighted covariate. That remains the default behavior. For
landscape metrics, however, the relationship between raster layers and model
covariates is no longer one-to-one.

For example, the same binary forest raster might be used to calculate:

- a distance-weighted proportion of forest cover,
- edge density within a fixed buffer, and
- edge density at an optimized scale.

Those are different covariates, even though they come from the same source
raster. The `msr_vars()` interface lets you name those derived covariates
explicitly.

---

# Load example data

This example uses the simulated count data and survey locations included with
`multiScaleR`. The data contain 100 spatial sample points, which is large enough
to demonstrate the workflow without making the vignette expensive to build.

```{r load-data}
data("landscape_counts")
dat <- landscape_counts

data("surv_pts")
pts <- vect(surv_pts)

land_rast <- rast(pkg_extdata("landscape.tif"))
```

The raster layer `land1` is a binary habitat layer. To demonstrate a categorical
landscape-diversity metric, we also create a simple three-class land-cover layer
from the continuous `land2` raster. In a real analysis, this would usually be a
mapped categorical land-cover product.

```{r build-landcover}
landcover <- classify(
  land_rast$land2,
  rcl = matrix(
    c(-Inf, -0.5, 1,
      -0.5,  0.5, 2,
       0.5,  Inf, 3),
    ncol = 3,
    byrow = TRUE
  )
)
names(landcover) <- "landcover"

metric_rasters <- c(land_rast$land1, landcover)
```

```{r plot-inputs, fig.cap = "***Source rasters used to derive model covariates.***"}
plot(metric_rasters)
plot(metric_rasters$land1, main = "land1 with survey points")
points(pts, pch = 19, cex = 0.6, col = "red")
```

---

# Define derived covariates

Use `msr_vars()` to define model covariates as transformations of source raster
layers. Here, `land1_prop` is optimized as a kernel-weighted covariate, while
`land1_ed_500` and `landcover_shdi_500` are fixed-radius landscape metrics
calculated within 500 m of each point.

```{r define-vars}
scale_vars <- msr_vars(
  land1_prop = kernel_var("land1"),
  land1_ed_500 = landscape_var("land1", metric = "ed", radius = 500),
  landcover_shdi_500 = landscape_var("landcover", metric = "shdi", radius = 500)
)

scale_vars
```

This is the key distinction:

- `land1` and `landcover` are source raster layers.
- `land1_prop`, `land1_ed_500`, and `landcover_shdi_500` are model covariates.
- Only `land1_prop` receives an optimized sigma parameter in this example.

---

# Prepare model data

Pass the covariate specification to `kernel_prep()`. The returned
`kernel_dat` data frame contains the derived covariates, not merely the original
raster layer names.

```{r kernel-prep}
kernel_inputs <- kernel_prep(
  pts = pts,
  raster_stack = metric_rasters,
  max_D = 1200,
  kernel = "gaussian",
  scale_vars = scale_vars,
  verbose = FALSE
)

head(kernel_inputs$kernel_dat)
```

Combine the derived covariates with the response data and fit the starting
model.

```{r fit-model}
df <- data.frame(dat, kernel_inputs$kernel_dat)

mod <- glm(
  counts ~ site + land1_prop + land1_ed_500 + landcover_shdi_500,
  family = poisson(),
  data = df
)
```

---

# Optimize the kernel covariate

Because the two landscape metrics were defined with fixed radii, only
`land1_prop` is optimized. This keeps the example fast while still showing how
kernel-scaled and landscape-metric covariates can be included in the same model.

```{r optimize}
opt <- multiScale_optim(
  fitted_mod = mod,
  kernel_inputs = kernel_inputs,
  verbose = FALSE
)
```

```{r summary}
summary(opt)
```

---

# Project derived covariates to rasters

`kernel_scale.raster()` also understands `scale_vars`. When a fitted
`multiScaleR` object is supplied, optimized covariates use their fitted scale
and fixed-radius landscape metrics use their specified radius.

```{r project, fig.cap = "***Projected derived covariates.***"}
projected <- kernel_scale.raster(
  raster_stack = metric_rasters,
  multiScaleR = opt,
  scale_center = TRUE,
  clamp = TRUE,
  verbose = FALSE
)

plot(projected)
```

The resulting raster names match the model covariates:

```{r projected-names}
names(projected)
```

---

# Optimizing landscape metric radii

To optimize the scale of a landscape metric, omit `radius` from
`landscape_var()`. For example:

```{r optimized-landscape-vars, eval = FALSE}
scale_vars_opt <- msr_vars(
  land1_prop = kernel_var("land1"),
  land1_ed = landscape_var("land1", metric = "ed"),
  landcover_shdi = landscape_var("landcover", metric = "shdi")
)
```

This is statistically meaningful, but it is more computationally expensive than
optimizing a kernel-weighted mean. At each candidate radius, the landscape metric
must be recalculated for every sample point. For moderate or large data sets,
consider first testing the workflow with a subset of points or using
precomputed results in reproducible documents.

---

# Practical guidance

Use fixed-radius landscape metrics when you have a biologically motivated buffer
distance or when the metric is expensive to recompute. Use optimized landscape
metric radii when the scale itself is a central question and computation time is
acceptable.

At present, optimized landscape metrics are not supported with
`kernel = "expow"` because the exponential-power shape parameter does not apply
cleanly to fixed-buffer landscape metrics.
