---
title: "Surface Texture Covariates"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Surface Texture Covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(multiScaleR)
library(terra)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4.25,
  warning = FALSE,
  message = FALSE
)

pkg_extdata <- function(...) {
  src_path <- normalizePath(file.path("..", "inst", "extdata", ...),
                            winslash = "/",
                            mustWork = FALSE)
  if (file.exists(src_path)) {
    return(src_path)
  }

  inst_path <- system.file("extdata", ..., package = "multiScaleR")
  if (nzchar(inst_path) && file.exists(inst_path)) {
    return(inst_path)
  }

  stop("Could not locate required extdata file: ", paste(..., collapse = "/"))
}
```

# What surface texture covariates measure

A kernel covariate in `multiScaleR` summarizes the *average* value of a
continuous raster in the neighborhood of a point. A surface texture covariate
summarizes something different: how *variable* that surface is within the
neighborhood. Two points can have the same average elevation, NDVI, or canopy
height, yet one can sit in smooth terrain and the other in rugged terrain.
Surface texture metrics capture that difference.

`surface_var()` derives these metrics from a continuous raster within a circular
buffer, in the same way that `landscape_var()` derives landscape metrics from a
categorical raster. The two functions are deliberately separate because they
expect opposite kinds of raster: `landscape_var()` requires integer class
codes, while `surface_var()` requires continuous values. If you have not used
the explicit covariate interface before, read
`vignette("landscape_metric_covariates", package = "multiScaleR")` first; this
vignette assumes you are comfortable with `msr_vars()`.

This release supports two surface texture metrics, both borrowed from the
surface-metrology literature and matched to the `geodiv` package so that
`geodiv` can be used to validate results.

## Surface metric reference

Each metric is computed on the continuous values within a circular buffer
centered on a sample point (or on each raster cell when projecting with
`kernel_scale.raster()`). In the formulas below, $z_i$ is a cell value in the
neighborhood, $\bar z$ is the mean of those values, and $n$ is the number of
finite cells.

| Code | Name | Definition | Reading |
|------|------|------------|---------|
| `sa` | Average roughness | $S_a = \frac{1}{n}\sum_i \lvert z_i - \bar z \rvert$ | Mean absolute deviation from the neighborhood mean. Robust to a few extreme cells. |
| `sq` | RMS roughness | $S_q = \sqrt{\frac{1}{n-1}\sum_i (z_i - \bar z)^2}$ | The sample standard deviation of the neighborhood. More sensitive to large deviations than `sa`. |

Both metrics are zero on a perfectly flat surface and grow as the surface
becomes more variable. They are reported in the units of the raster itself: if
`land3` ranges from 0 to 1, a roughness of 0.10 means the values within the
buffer deviate from their local mean by about a tenth of that range. `sa` and
`sq` measure the same underlying idea (local spread) and are usually
correlated; `sq` weights large deviations more heavily because it squares them,
so it responds more strongly to a few unusual cells.

## A note on scale

A surface texture covariate behaves differently from a kernel-weighted mean as
you change the radius. A larger radius does not simply smooth the same quantity
over a wider area; it changes what is being measured. A small radius asks how
rough the surface is in the immediate vicinity of a point (fine-grain texture).
A large radius asks how variable the surface is across a broader extent, which
folds in any large-scale gradient or patchiness. Keep this in mind when
interpreting an optimized radius: the estimated scale is the neighborhood size
at which within-neighborhood variability best explains the response, not the
scale of an averaging window.

---

# Load example data

This example reuses the simulated count data, survey locations, and landscape
raster included with `multiScaleR`. The data contain 100 spatial sample points,
which is enough to demonstrate the workflow without making the vignette
expensive to build.

```{r load-data}
data("landscape_counts")
dat <- landscape_counts

data("surv_pts")
pts <- vect(surv_pts)

land_rast <- rast(pkg_extdata("landscape.tif"))
```

The raster stack contains two continuous layers suited to surface texture
metrics. `land3` is a smooth, highly autocorrelated surface (think elevation or
canopy height), so its roughness varies meaningfully from place to place.
`land2` is a noisier, weakly autocorrelated surface. We also keep `land2`
because the example counts were simulated to respond to its kernel-weighted
mean, which gives us a covariate with a genuine, interpretable scale of effect.

```{r plot-inputs, fig.cap = "***Continuous source surfaces. land3 (right) is smooth; land2 (left) is noisy.***"}
surfaces <- c(land_rast$land2, land_rast$land3)
plot(surfaces)
plot(land_rast$land3, main = "land3 with survey points")
points(pts, pch = 19, cex = 0.6, col = "red")
```

**How to read these plots:** Where `land3` transitions sharply between low and
high values, a buffer placed on the transition contains a wide spread of values
and will have high roughness. Where `land3` is locally flat (a broad light or
dark patch), a buffer contains similar values and roughness is low. The
roughness surfaces produced later in this vignette make that pattern explicit.

---

# Define derived covariates

Use `msr_vars()` to name the model covariates. Here `land2_mean` is an
optimized, kernel-weighted mean, while `land3_sq_500` and `land3_sa_500` are
fixed-radius surface texture metrics computed within 500 m of each point.
Computing the surface metrics at a fixed radius keeps this example fast: they
are calculated once and are not re-evaluated during optimization.

```{r define-vars}
scale_vars <- msr_vars(
  land2_mean   = kernel_var("land2"),
  land3_sq_500 = surface_var("land3", metric = "sq", radius = 500),
  land3_sa_500 = surface_var("land3", metric = "sa", radius = 500)
)

scale_vars
```

The distinction is the same as for landscape metrics:

- `land2` and `land3` are source raster layers.
- `land2_mean`, `land3_sq_500`, and `land3_sa_500` are model covariates.
- Only `land2_mean` receives an optimized sigma in this example.

---

# Prepare model data and fit

Pass the specification to `kernel_prep()`. The returned `kernel_dat` data frame
holds the derived covariates, already centered and scaled to unit variance.

```{r kernel-prep}
kernel_inputs <- kernel_prep(
  pts = pts,
  raster_stack = land_rast,
  max_D = 1500,
  kernel = "gaussian",
  scale_vars = scale_vars,
  verbose = FALSE
)

head(kernel_inputs$kernel_dat)
```

Combine the covariates with the response and fit the starting model.

```{r fit-model}
df <- data.frame(dat, kernel_inputs$kernel_dat)

mod <- glm(
  counts ~ site + land2_mean + land3_sq_500 + land3_sa_500,
  family = poisson(),
  data = df
)
```

---

# Optimize and interpret

Only `land2_mean` was left free, so optimization estimates a single scale
parameter while the two surface metrics stay fixed at 500 m. This keeps the
example fast and mirrors a common analysis pattern: optimize the covariate
whose scale is the scientific question, and hold the others at a biologically
motivated buffer.

```{r optimize}
opt <- multiScale_optim(
  fitted_mod = mod,
  kernel_inputs = kernel_inputs,
  verbose = FALSE
)
```

```{r summary}
summary(opt)
```

**How to read the `summary()` output:** The scale section reports the estimated
sigma for `land2_mean`, with its standard error and 95% confidence interval, in
map units (meters here). Below it is the *effective distance*, the radius that
captures a stated proportion (default 90%) of the Gaussian kernel weight. That
converts sigma into a distance you can interpret on the ground. The two
fixed-radius surface metrics do not appear in the scale section because their
radius was specified rather than estimated; they enter the model only as
ordinary covariates.

The model coefficients are on the Poisson log (link) scale, so they must be
back-transformed to be interpretable. Because every covariate was centered and
scaled, each coefficient describes the effect of a one standard deviation
change in that covariate. Exponentiating gives the multiplicative change in
expected count.

```{r coef-interpret}
cf <- coef(summary(opt$opt_mod))
## Multiplicative change in expected count per 1 SD increase in each covariate
round(exp(cf[, "Estimate"]), 3)
```

For example, an `exp()` value of 1.20 for a surface metric would mean that a one
standard deviation increase in roughness is associated with a 20 percent higher
expected count, holding the other covariates constant. Note an important caveat:
these example counts were simulated to respond to kernel-weighted means of
`land1` and `land2`, not to surface roughness. The recovered scale for
`land2_mean` is therefore the meaningful quantity to interpret here, and the
roughness coefficients are best read as a demonstration of the mechanics. In
your own analysis, a surface texture coefficient is interpreted exactly like any
other scaled predictor.

---

# Project covariates to rasters

`kernel_scale.raster()` understands `surface_var()` specifications just as it
does kernel and landscape covariates. When a fitted `multiScaleR` object is
supplied, the optimized covariate uses its fitted scale and the fixed-radius
surface metrics use their specified radius. Setting `scale_center = TRUE`
returns the projected covariates on the same centered and scaled footing as the
model, ready for `terra::predict()`.

```{r project, fig.cap = "***Projected model covariates at their fitted or fixed scales.***", fig.height = 6}
projected <- kernel_scale.raster(
  raster_stack = land_rast,
  multiScaleR = opt,
  scale_center = TRUE,
  clamp = TRUE,
  verbose = FALSE
)

plot(projected)
names(projected)
```

**How to read these plots:** `land2_mean` is a smoothed average surface, so it
looks like a blurred version of the source layer. The roughness layers
(`land3_sq_500`, `land3_sa_500`) look different: they are bright where `land3`
changes rapidly within 500 m and dark where it is locally uniform. Roughness
maps highlight transitions and texture, not high or low values. The stack also
includes a constant `site` layer filled with zeros: because `site` is a
field-measured covariate with no raster source, `kernel_scale.raster()` appends
a placeholder so the output can be passed directly to `terra::predict()`.
Replace it with a real surface if you want to predict at a non-reference value
of `site`.

---

# Roughness surfaces without a fitted model

You do not need a fitted model to produce a roughness surface. Pass a
`scale_vars` object in which every covariate has a fixed `radius`, and
`kernel_scale.raster()` returns the metric surfaces directly. This is useful for
exploring how texture varies across a landscape, or for comparing scales before
committing to a model.

```{r standalone-multiscale, fig.cap = "***RMS roughness (sq) of land3 at three radii. Larger radii fold in broader-scale variation.***", fig.height = 3.5}
sq_multi <- kernel_scale.raster(
  raster_stack = land_rast["land3"],
  scale_vars = msr_vars(
    sq_200 = surface_var("land3", metric = "sq", radius = 200),
    sq_500 = surface_var("land3", metric = "sq", radius = 500),
    sq_900 = surface_var("land3", metric = "sq", radius = 900)
  ),
  verbose = FALSE
)

plot(sq_multi, main = c("sq - 200 m", "sq - 500 m", "sq - 900 m"))
```

**How to read these plots:** At 200 m the surface picks out fine, local
texture, so roughness is patchy and concentrated near sharp transitions. At
900 m each buffer spans more of the underlying gradient, so roughness is
smoother and generally higher. This is the practical face of the scale caution
from the introduction: the radius changes the quantity, not just its
resolution.

The `sa` and `sq` metrics emphasize roughness slightly differently. Comparing
them at a single radius shows where a few extreme cells (which `sq` weights more
heavily) drive the signal.

```{r sa-vs-sq, fig.cap = "***Average roughness (sa) versus RMS roughness (sq) of land3 at 500 m.***", fig.height = 3.5}
sa_sq <- kernel_scale.raster(
  raster_stack = land_rast["land3"],
  scale_vars = msr_vars(
    sa_500 = surface_var("land3", metric = "sa", radius = 500),
    sq_500 = surface_var("land3", metric = "sq", radius = 500)
  ),
  verbose = FALSE
)

plot(sa_sq, main = c("sa - 500 m", "sq - 500 m"))
```

Projection of `sq` uses fast Fourier transform convolution, and projection of
`sa` uses a compiled focal pass. Both reproduce the per-point calculation
exactly; `sq` is the faster of the two on very large rasters.

---

# Optimizing the scale of roughness

To estimate the scale of a surface texture covariate rather than fix it, omit
`radius` from `surface_var()`. The covariate then receives its own sigma, which
`multiScale_optim()` estimates alongside the model.

```{r optimized-surface-vars, eval = FALSE}
scale_vars_opt <- msr_vars(
  land2_mean = kernel_var("land2"),
  land3_sq   = surface_var("land3", metric = "sq")
)

kernel_inputs_opt <- kernel_prep(
  pts = pts,
  raster_stack = land_rast,
  max_D = 1500,
  scale_vars = scale_vars_opt,
  verbose = FALSE
)
```

Optimizing a surface texture radius is well defined, but it costs more than
optimizing a kernel-weighted mean: at each candidate radius the metric is
recomputed for every sample point, since (unlike a kernel mean) it cannot be
read off a precomputed distance binning. Surface metrics are continuous in the
radius, so they tend to optimize smoothly. Start with fixed radii to confirm
the workflow, then optimize once the model structure is settled. Optimized
surface metrics are not currently supported with `kernel = "expow"`, because the
exponential-power shape parameter does not apply to a fixed-buffer metric.

---

# Practical guidance

- Use a surface texture covariate when you expect the *variability* of a
  continuous surface, rather than its average, to matter ecologically: terrain
  ruggedness, vegetation-structure heterogeneity, thermal patchiness, and
  similar.
- Choose `sq` when large local deviations are ecologically important, and `sa`
  when you want a measure that is less sensitive to a few extreme cells.
- A surface metric and a kernel-weighted mean from the same source layer answer
  different questions and can both be included; check their correlation before
  interpreting either in isolation.
- Use a fixed radius when you have a biologically motivated buffer or want a
  fast model; optimize the radius when the scale of heterogeneity is itself the
  question.

---

# Quick reference

```{r quick-reference, eval = FALSE}
## 1. Define covariates: an optimized kernel mean plus fixed-radius roughness
scale_vars <- msr_vars(
  land2_mean   = kernel_var("land2"),
  land3_sq_500 = surface_var("land3", metric = "sq", radius = 500),
  land3_sa_500 = surface_var("land3", metric = "sa", radius = 500)
)

## 2. Extract and prepare the covariates around each point
kernel_inputs <- kernel_prep(pts, land_rast, max_D = 1500,
                             scale_vars = scale_vars, verbose = FALSE)

## 3. Fit the starting model
df  <- data.frame(dat, kernel_inputs$kernel_dat)
mod <- glm(counts ~ site + land2_mean + land3_sq_500 + land3_sa_500,
           family = poisson(), data = df)

## 4. Optimize the free scale parameter(s)
opt <- multiScale_optim(fitted_mod = mod, kernel_inputs = kernel_inputs)
summary(opt)                              # sigma + effective distance, on map units

## 5. Project covariates for prediction (centered and scaled like the model)
projected <- kernel_scale.raster(land_rast, multiScaleR = opt,
                                 scale_center = TRUE, clamp = TRUE)

## Standalone roughness surface, no model required
sq_500 <- kernel_scale.raster(
  land_rast["land3"],
  scale_vars = msr_vars(sq_500 = surface_var("land3", metric = "sq", radius = 500))
)
```

## Summary of key functions

| Function | Purpose |
|----------|---------|
| `surface_var()` | Define a continuous surface texture covariate (`sa` or `sq`). |
| `msr_vars()` | Collect covariate specifications into one object. |
| `kernel_prep()` | Extract and prepare covariates around each point. |
| `multiScale_optim()` | Estimate the scale of effect for optimized covariates. |
| `kernel_scale.raster()` | Project covariates to rasters, with or without a fitted model. |

## See also

- `vignette("landscape_metric_covariates", package = "multiScaleR")` for
  categorical landscape metrics and the `msr_vars()` interface.
- `vignette("multiScaleR_Guide", package = "multiScaleR")` for the full
  modeling workflow and model interpretation.
- `vignette("spatial_projection_clamping", package = "multiScaleR")` for
  projecting fitted models to prediction surfaces.
